---
title: "data.table"
author: Gregor and Selina
date: Wednesday, May 2, 2018
output: ioslides_presentation
widescreen: true
css: style.css
---


##What is data.table?
- package released in 2008 by Matt Dowle  

__main goals__:  

- reduced programming time --> more natural syntax  
    - less variable name repetition  
    - most things can be done in a single line of code  
- reduced computing time  
    - fast aggreagation  
    - update by reference  

##Why to use it?

- faster read, write & data manipulation  
    - works with shallow copies & pointers (column names) not deep copies
```{r}
library(nycflights13)
library(plyr)
library(data.table)  

dt<-data.table(do.call("rbind", replicate(5, flights, simplify = FALSE)))
system.time(ddply(dt, .(origin, carrier, month), summarise, delay = mean(dep_delay)))
system.time(dt[,mean(dep_delay),.(origin,carrier,month)])
```


##data.table vs. data.frame
- data.table inherits from data.frame  
    --> every data.table is also a data.frame and can be used in the same way  
- accepted as data.frame by other packages  


```{r}
dt2 <- data.table(A = c(1L,2L), B = LETTERS[1:3], C = round(rnorm(4),4), D = 1:12)
class(dt2)
head(dt2)

```



##Basic Syntax of data.table

__dt[<span style="color:red">i</span>, <span style="color:green">j</span>, <span style="color:blue">by</span>]__

"Take dt, subset rows using __<span style="color:red">i</span>__, then calculate __<span style="color:green">j</span>__ grouped by __<span style="color:blue">by</span>__"


##Subsetting rows using i
```{r}
dt2[1:2] #select 1st and 2nd row (same as dt2[1:2,])
dt2[10:(.N-1)] #select 9th to penultimate row
dt2[B %in% c("A","C")] #select all rows that have value A or C in column B

```
##Manipulating on columns j:
```{r}
dt2[1:2,B] # returns a vector
dt2[1:2,.(B)] # returns a data.table
dt2[1:2,.(B,C)] #returns a data.table
```


##Manipulating on columns j:

- .() is short for list()
- any expression allowed in j (even plot)
- j is evaluated within the scope of dt --> no dt$colname=...
- column names are treated like variables --> no ""


##Summarizing data.tables{.smaller}
```{r}
dt <- data.table(flights)
```
delay statistics:
```{r}
dt[,.(mean_del = mean(arr_delay, na.rm=T), sd_del = sd(arr_delay, na.rm=T))]
```
returns a new data table:
```{r}
summ_tab <- dt[,.(mean_del = mean(arr_delay, na.rm=T), sd_del = sd(arr_delay, na.rm=T))]
str(summ_tab)
```


##Summarizing data.tables cont'd{.smaller}

more sophisticated delay statistics:

```{r}
summ_tab <- dt[,.(mean_del = mean(arr_delay, na.rm=T),
                  sd_del = sd(arr_delay, na.rm=T)),
               by = .(carrier, month, origin)]
str(summ_tab)
```

remember:
```{r, eval=F}
###<b>
dt[i, j, by]
###</b>
```


##Updating columns
let's keep it simple
```{r}
dt[, arr_delay := round(arr_delay, 0)]

dt[, c("arr_delay", "dep_delay")
   := list(round(arr_delay, 0),
           round(dep_delay, 0))]
```
let's keep it even simpler
```{r}
dt[, day := NULL]

dt[, c("year", "arr_time") := NULL]
```




##Keys{.smaller}
the master coulmns of data.tables

```{r}
setkey(dt, carrier)

head(dt["B6"])
```



##Keys cont'd{.smaller}
```{r}
sub_tab <- dt[c("UA", "AA")]

dt["B6", mult="first"]
```

```{r}
dt[c("UA", "AA"), mean(arr_delay, na.rm = T)]
dt[c("UA", "AA"), mean(arr_delay, na.rm = T), by = carrier]
```


##Chaining operations and queries
rept: summarising ```by```
```{r}
summ_tab <- dt[,.(mean_del = mean(arr_delay, na.rm=T)),
               by = .(carrier, month, origin)]
```


"piping" another operation: select only the major delays
```{r hilang = 'sas'}
summ_tab <- dt[,.(mean_del = mean(arr_delay, na.rm=T)),
               by = .(carrier, month, origin)][mean_del > 2]
```




##Read
```{r, eval = F}
fread()
```
- similar to read.table but much faster
- controls (sep, colClasses, ...) auto-detected
- can read directly from web
```{r, eval = F}
mydata =  
  fread("https://github.com/arunsrinivasan/satrdays-workshop/raw/master/flights_2014.csv")
```

##Read cont'd
speeding things up:
```{r, eval=F}
library(paarallel)

file.list <- list.files(pattern = "\\.txt$")
 
cores <- detectCores()
 
cl <- makeCluster(cores)

dt <- do.call(rbind, parLapply(cl, X=file.list,
              fun = function(x){
                  return(data.table::fread(x, header = TRUE, fill = T))
                  )})
```


##Write
```{r,eval=F}
fwrite()
```

- as write.csv but faster (2 s vs 1 min)
- enabled with parallel computing

##


and always remember your best friend 
```
?data.table
```
